---
title:  |   
  | Using hierarchical joint models to study reproductive interactions in plant communities:  
  | Individual-level fitness data
author: "Øystein H. Opedal & Stein Joar Hegland"
date: "9 May 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/data/HMSC_poll_inter/")
library(Hmsc)
library(corrplot)
library(knitr)
```

Contact: oystein.opedal@helsinki.fi

# 1. Introduction
Here, we demonstrate how to set up a HMSC analysis of fitness data (e.g. seed set) for 5 species cooccurring on 50 plots. Two traits (z~1~ and z~2~) are measured for each individual plant. We will use these data to estimate residual fitness correlations among species, i.e. how the fitness of each species in each plot affects the fitness of other species when growing in those plots.

Furthermore, we will use the trait and fitness data to jointly estimate phenotypic selection gradients for the two traits in each species using the method of Lande and Arnold 1983 (Evolution).

For technical details about the HMSC model, see Ovaskainen et al. 2017 (Ecology Letters) and Tikhonov et al. 2019 (bioRxiv).

# 2. Preparing data

## Load the study design file

The file `SimulatedStudyDesign.csv` includes trait measurements for 5 species occurring on 50 plots, and an indicator of which species each row corresponds to. There is more than one individual of each species in some plots.

```{r I, message=F, warning=F, cache=T}
library(Hmsc)
library(corrplot)

alldat = read.csv("SimulatedStudyDesign.csv")
head(alldat)
```

## Prepare the trait data

Trait data and other covariates are provided as a dataframe (`XData`), and a formula describing the regression equation (`XFormula`). The formula follows standard `R` syntax. HMSC does not currently allow missing values in `XData`, hence individuals with missing values must be excluded, or the missing values replaced with means or other gap-filling techniques.

```{r Ib, message=F, warning=F}
XData = alldat[,2:3]
head(XData)

XFormula = ~ z1 + z2
```

## Prepare the `studyDesign` dataframe

The study design is provided as a dataframe containing factors indicating the hierarchical structure of the data. Here, individuals belong to one of the 50 plots. Additional random effects can be included by adding columns to the `studyDesign` dataframe.

```{r Ic, message=F, warning=F}
studyDesign = data.frame(plot = alldat$plot)
studyDesign$plot = as.factor(studyDesign$plot)
head(studyDesign)
```

## Generate fitness data
We simulate fitness data for each individual as a function of their traits z~1~ and z~2~. We let the fitness of Species 1 in each plot affect the fitness of Species 2 and 3 in that plot positively, and the fitness of Species 4 and 5 in that plot negatively.

Note that each line in the **Y** matrix correspond to one individual of one species, and all other species are set to NA.

```{r II}
Y = alldat[,4:8]

alpha = c(25, 7, 1, 5, 40)
beta_z1 = c(-1.8, 0.5, -1.2, 0.4, -1.1)
beta_z2 = c(1.4, -0.3, 0.7, 0.4, 0.2)
sp1eff = c(0, 1, 1, -1, -1)
set.seed(1)

Y[,1] = alpha[1] + beta_z1[1]*XData$z1 + beta_z2[1]*XData$z2 + rnorm(nrow(Y), 0, 1) + 
        rnorm(50,0,2)[studyDesign$plot]

sp1plots = tapply(Y[,1], studyDesign$plot, mean, na.rm=T)

for(s in 2:5){
Y[,s] = alpha[s] + beta_z1[s]*XData$z1 + beta_z2[s]*XData$z2 + rnorm(nrow(Y), 0, 1) + 
        rnorm(50,0,.5)[studyDesign$plot] + sp1eff[s]*sp1plots[studyDesign$plot]
}

Y = Y*as.numeric(alldat[,4:8]>0)
apply(Y, 2, range, na.rm=T)

Y[c(1:5, 363:367),]
```

## Convert fitness values to relative fitness

To estimate selection gradients, we divide the fitness values by the mean fitness for each species.

```{r IIb}
Y = apply(Y, 2, function(x) x/mean(x, na.rm=T))
```

# 3. Setup the HMSC model

## Define a HMSC random level for plots

A plot-level random effect will allow us to assess residual correlations among species' fitness values in a given plot. Multiple random levels can be set up. For spatially explicit data, the argument `units` is replaced by providing a matrix of x and y coordinates with the argument `sData`. 

```{r III}
rL1 = HmscRandomLevel(units = unique(studyDesign$plot))
```

## Setup the HMSC model

At this stage we also set the error distribution of the analyses, here "normal" for Gaussian errors.

```{r IV}
m = Hmsc(Y=as.matrix(Y), XData = XData,  XFormula = XFormula,
         dist = "normal", studyDesign = studyDesign, ranLevels = list(plot=rL1))
```

# 4. Sample the posterior distribution

As with all MCMC-based Bayesian analyses, the posterior needs to be sampled until convergence. A good strategy is to start with a small number of iterations, and then increase the number of iterations until the results converge. We run 2 replicate MCMC chains to assess whether these converge. 

```{r V, cache=T}
samples = 1000
thin = 5
adaptNf = .4*(samples*thin)
transient = .5*(samples*thin)
nChains = 2

start = Sys.time()
m = sampleMcmc(m, samples = samples, thin = thin, transient = transient, 
             verbose = F, adaptNf = rep(adaptNf, m$nr), nChains = nChains)
Sys.time()-start
```

# 5. Assessing model convergence

To assess chain convergence, we can look at posterior trace plots, effective sample sizes, and potential scale reduction factors. In this case, all of these indicate convergence of the sampling, because there is no trend in the posterior trace plots, the two chains overlap, the effective sample sizes are close to the number of samples, and the potential scale reduction factors are close to 1.

## Extract posterior and convert to `Coda` object

```{r VI, cache=F, fig.height=6, fig.width=6, fig.cap="\\label{fig: VI} Posterior trace plot of the first three beta parameters"}
post = convertToCodaObject(m)  
plot(post$Beta[,1:3])
```

## Effective sample size for beta parameters

```{r VII, cache=F}
summary(effectiveSize(post$Beta))
```

\pagebreak

## Potential scale reduction factors for beta parameters

```{r VIII, cache=F}
summary(gelman.diag(post$Beta)$psrf)
```

# 5. Evaluate model fit

HMSC comes with tools for computing measures of model fit for each species

```{r IX, cache=F}
predY = computePredictedValues(m)
MF = evaluateModelFit(m, predY)
MF
```

# 6. Assessing parameter estimates

## Compute and plot variance partitioning

HMSC also comes with tools for performing variance component analyses and plotting the results. Because we have many `NA`'s in the **Y** matrix, we use the `na.ignore = TRUE` option to compute the variances and covariances of the traits only for those sites where the species is present.
  
```{r X, cache=F, message=F, warning=F, fig.width=8, fig.cap="\\label{fig: X} Variance partitioning for each species"}
group = c(1, 1, 2)
groupnames = m$covNames[-1]
VP = computeVariancePartitioning(m, group = group, groupnames = groupnames, na.ignore = TRUE)
par(mar=c(4,4,2,12), xpd=T)
plotVariancePartitioning(m, VP = VP, args.legend=list(x=9.3, y=1, bty="n"))
```

\pagebreak

## Extract and plot beta parameters

The regression coefficients for the fixed part of the HMSC model can be extracted with the `getPostEstimate` function and visualised with the `plotBeta` function.

```{r XI, cache=F, fig.height=4, fig.width=6, fig.cap="\\label{fig: XI} Heatmap of beta parameters (raw selection gradients)"}
mbeta = getPostEstimate(m, "Beta")
mbeta
par(mar = c(0,0,0,0))
plotBeta(m, post = mbeta, "Mean", covOrder = "Vector", covVector = c(2:3), supportLevel=0)
```

## Extract selection gradients

To compare the strength of selection across traits and species, selection gradients are typically standardized either by trait means or standard deviations. Mean-standardization yields a measure of proportional change in fitness per proportional change in the trait (i.e. an elasticity), while SD-standardization yields a measure of proportional change in fitness per standard deviation change in the trait (sometimes called selection intensity or *i*). 

```{r XII, cache=F, echo=F}
z1Means = colMeans(XData$z1*as.matrix(alldat[,4:8]),na.rm=T)
z2Means = colMeans(XData$z2*as.matrix(alldat[,4:8]),na.rm=T)
z1SD = apply(XData$z1*as.matrix(alldat[,4:8]), 2, sd, na.rm=T)
z2SD = apply(XData$z2*as.matrix(alldat[,4:8]), 2, sd, na.rm=T)

betamyz1 = mbeta$mean[2,]*z1Means
betamyz2 = mbeta$mean[3,]*z2Means
betavarz1 = mbeta$mean[2,]*z1SD
betavarz2 = mbeta$mean[3,]*z2SD

sumstats = matrix(NA,nrow=10, ncol=6)
sumstats[,1] = rep(c("Mean", "SD", "Beta", "Beta_mean", "Beta_var"),2)
sumstats[,2:6] = round(rbind(z1Means,z1SD,mbeta$mean[2,],betamyz1, betavarz1, 
                           z2Means, z2SD, mbeta$mean[3,], betamyz2,betavarz2),2)
rownames(sumstats) = c("z1",rep("",4),"z2",rep("",4))
colnames(sumstats) = c("",paste("Species",1:5))
kable(sumstats, caption="Means, standard deviations, and selection gradients for each species")
```

\pagebreak

## Compute and plot species associations

Residual correlations at each random level (here plot) can be extracted with the `computeAssociations`function. The estimated associations match those assumed when simulating the data.

```{r XIII, cache=F, fig.cap="\\label{fig:XIII} Residual fitness correlations"}
OmegaCor = computeAssociations(m)
supportLevel = 0.75

toPlot = ((OmegaCor[[1]]$support>supportLevel) + 
            (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean

corrplot(toPlot, type="lower", tl.cex=.7,tl.col="black",tl.srt=45,method = "color", 
         col=colorRampPalette(c("blue","white", "red"))(200), mar=c(0,0,0,1)) 
```
